Cohesiveness in microbial community coalescence

### 2021-05-11 draft/draft.tex
finished draft, time to run through alvaro

% title, authors, affiliations
\title{Provisional Title}
\author[1,2,3]{Authors}
\author[1,2,$\dagger$]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology,
Yale University, New Haven, CT, USA}
\affil[2]{Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[3]{Other affiliations...}
\affil[$\dagger$]{To whom correspondence should be addressed: \normalfont alvaro.sanchez@yale.edu}
\date{}

@@

Microbial communities often invade one another.
This has been observed, for instance, 
in river courses where terrestrial microbial communities mix with aquatic microorganisms
\cite{Mansour2018,Luo2020,Vass2021}
or in soil communities being invaded as a result of
tillage and outplanting \cite{Rillig2016} or by
aerially dispersed bacteria and funghi \cite{Evans2019}.
The human digestive system can get invaded several times a day by the microbial consortia
that reside on the ingested food, and the skin microbiota is also subject to
invasions when making contact with environmental sources of microbes \cite{Vandegrift2019}.

The phenomenon by which entire microbiomes invade one another has been termed
\textit{community coalescence} \cite{Rillig2015}.
Ecologists have long contemplated the idea that interactions between multiple co-invading species
can produce correlated invasional outcomes
\cite{Gilpin1994,Grosholz2005,Green2011,Livingston2013,Rillig2015,OLoughlin2017,Castledine2020}.
One example is the hypothesis known as \textit{invasional meltdown}, which proposes that
positive interactions between co-invading species can enhance their invasive success and
facilitate future invasions \cite{Simberloff1999,Simberloff2006,Gurevitch2006}.
However, and
in spite of its clear potential importance, the role of coalescence in microbiome assembly is
only beginning to be addressed and little is known about the mechanisms that govern it and its
potential implications.
Early mathematical models of community-community invasions \cite{Gilpin1994,Toquenaga1997}
as well as more recent work
\cite{Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021}
suggest that high-order invasion effects are common
during community coalescence. Communities that have a previous history of coexistence may exhibit an
emergent ``cohesiveness'' which produces correlated invasional outcomes among species from the
same community \cite{Livingston2013,Sierocinski2017}.
The situation where ecological partners in the invading community recruit each other into the final
coalesced community has been called \textit{ecological co-selection} \cite{Rillig2017,Sierocinski2017}.

The mechanisms of ecological co-selection during community coalescence are still poorly understood.
Do a few key species recruit everyone else, or are collective interactions among all species
(including the rarer members of the community) relevant for coalescence outcomes?
While it is reasonable to expect species with larger population sizes to have a proportionally oversized effect,
natural communities tend to be highly diverse \cite{Louca2016}
and
the role played by the less abundant species has long been subject to debate \cite{Winfree2015}.
Laboratory cultures have also been found to contain uneven distributions of multiple strains that feed
off the metabolic secretions of the dominant species \cite{Rosenzweig1994,Goldford2018}.
The fate of these sub-dominant taxa may be dependent on the invasional success of their dominant
species, or, alternatively, the dominant itself may owe its dominance (at least in part) to cross-feeding
or other forms of facilitation from the rarer members of the population.
These scenarios would give rise to ``top-down'' or ``bottom-up'' community cohesiveness, respectively.
Either of these forms of co-selection could, in principle, be positive (recruitment) or negative (antagonism),
as illustrated in \figref[e]{fig1}.
Which of these situations are typically found in nature?
Addressing this question has been experimentally challenging in the past \cite{Rillig2017,Sierocinski2017}.

In previous work, we have shown that a large amount of soil and plant microbiomes can be cultured
\textit{ex situ} in synthetic minimal environments with a single supplied limiting resource under serial
growth-dilution cycles \cite{Goldford2018} (\figref[a-b]{fig1}). Under these conditions, environmental
microbiomes spontaneously re-assemble into complex multi-species communities sustained by dense
cross-feeding facilitation networks \cite{Goldford2018}. In addition,
and just like in natural assemblies,
species abundance distributions
in these communities are generally long-tailed and uneven (\figref[d]{fig1} and \figref{figS1}),
with the dominant (most abundant) species typically comprising most of the biomass ($\text{median}=46\%$,
\figref{figS1}). Because these communities are easy to manipulate and grow in high throughput, and
are largely made up by culturable members, they represent good test cases to investigate ecological
co-selection during community coalescence. Here we focus on the dominants and ask whether they can
co-select or be co-selected by the sub-dominant species in their communities (henceforth referred to as
their \textit{cohorts}, \figref[c]{fig1}).
Our results confirm that positive top-down co-selection is common, but its effects are weak.
In contrast, bottom-up co-selection can be very strong, and positive co-selection is far more common
than negative co-selection.
We then turn to a microbial consumer-resource model (microCRM)
\cite{Goldford2018,Marsland2019,Marsland2020}
that is able to capture the dynamics of microbial communities dominated by metabolic interactions,
as is the case for the ones assembled in our experimental conditions \cite{Goldford2018}.
We show that the empirically observed trends in ecological co-selection are reproduced
with minimal model assumptions, and that
tuning the complexity of the metabolic interactions in our \textit{in silico} communities can modulate
the recurrence of top-down or bottom-up co-selection.
Our findings indicate that collective interactions play an important role at dictating community structure
during coalescence.

@@

We collected eight natural microbiomes from different soil and plant environmental samples
(\figref[a]{fig1})
and used them to inoculate our synthetic communities, which were stabilized 
in serial batch-culture biorreactors for 84 generations in synthetic minimal media containing either
glutamine or citrate as the only supplied carbon source
(\figref[b]{fig1}, \methodsref{community-assembly}).
We isolated the dominant species of every community
(\methodsref{dominants})
and identified them by Sanger-sequencing their 16S rRNA gene
(\methodsref{sequencing}), which correctly matched the dominant
Exact Sequence Variant (ESV) \cite{Callahan2016,Callahan2017} found through
community-level 16S illumina sequencing
(\figref{figS1}).
These dominants remained at high frequency after seven additional transfers
with the exception of two of the citrate communities and one of the glutamine
communities (where the dominants were presumably a transiently dominating species)
that were excluded from further analysis
(\figref{figS1}).
Similarly, pairs of communities where the dominants shared a same 16S sequence
and had similar colony morphology were excluded
(\figref{figS1}).

\subsubsection*{The strength of top-down co-selection is dependent on the primary supplied resource}

If communities being coalesced were highly cohesive from the top-down,
the dominant species would co-select the rarer members of its community during coalescence
(\figref[e]{fig1}, left panels).
In this scenario, we would expect the outcome of community coalescence to be predicted by
which of the two dominants is most competitive in pairwise competition.
Coalescence outcomes can be quantified by the similarity between the coalesced and the
invasive communities
(\methodsref{metrics}).
To test this hypothesis, we performed all pairwise competitions between dominant species
in glutamine and citrate environments by mixing them 1:1 on their native media and propagating
the cultures for seven serial transfers, roughly 42 generations
(\methodsref{competitions}).
We then performed all possible pairwise community coalescence experiments by mixing
equal volumes of the communities and propagating the resulting cultures for seven extra
transfers
(\figref[f]{fig1}).
The frequencies of all species in both community-community and dominant-dominant competitions
were determined by 16S illumina sequencing
(\methodsref{sequencing}).

We found that, for communities assembled in the glutamine environment,
the pairwise competitive ability of an invasive dominant
is only weakly predictive of the performance of the invasive community in coalescence
(as quantified by the relative Bray-Curtis similarity between the coalesced and invasive communities),
but this correlation was stronger for the citrate communities
(\figref[a]{fig2}, $R^2=0.15$, $p<0.05$ for glutamine and $R^2=0.57$, $p<10^{-4}$ for citrate).
Alternative quantifications of community distance yield similar results,
with weaker effects when the metric used
accounts only for the presence/absence of specific species and not for their relative abundance in
the communities
(\figref{figS2}).
All these metrics include the presence of the dominant species themselves. To better disentangle the effect
that these dominants have on the other member of their communities, we repeated the analysis this time
excluding the dominant species from the compositional data, finding that our results still hold
(\figref{figS3}).
Furthermore, we observed that, in the glutamine communities,
the relative frequency of a dominant against another in head-to-head pairwise competition
is barely predictive of its relative frequency against that same other dominant when the
cohorts are present too, i.e. after coalescence
($R^2=0.04$, $p>0.05$, \figref{figS4}).
On the other hand, pairwise competition between dominants of the citrate communities
is highly correlated with the relative abundance of said dominants in community coalescence
 ($R^2=0.83$, $p<10^{-8}$, \figref{figS4}).

Together, these results suggest that the strength of top-down co-selection depends on
the primary resource supplied to the coalesced communities.
Communities assembled with citrate as the primary supplied resource
display a strong degree of top-down co-selection,
with the fates of the sub-dominant species determined to a large extent by dominant-dominant
pairwise competition. This competition is, in turn, only weakly affected by the presence of the cohorts.
For glutamine communities, although some degree of top-down co-selection is consistent with our data,
the cohorts do not appear to be passively responding to their dominants.
Instead, evidence suggests that the cohorts might be playing an active role in community coalescence.
This finding led us to investigate the potential role of bottom-up ecological co-selection
(\figref[e]{fig1}, right panels), i.e. whether in some cases the dominants may be co-selected
for or against by their cohorts.

@@

\subsubsection*{Dominant-dominant and community-community competitions}
\label{methods:competitions}

@@

The sequencing protocol was identical to that described in \cite{Goldford2018}.
Community samples were collected by spinning down at 3500rpm for 25min
in a bench-top centrifuge at room temperature;
cell pellets were stored at $-80^\circ$C before processing.
To maximize Gram-positive bacteria cell wall lysis,
the cell pellets were re-suspended and incubated at $37^\circ$C for 30min
in enzymatic lysis buffer (20mM Tris-HCl, 2mM sodium EDTA, 1.2\% Triton X-100)
and 20mg/mL of lysozyme from chicken egg white (Sigma-Aldrich).
After cell lysis, the DNA extraction and purification was performed using the
DNeasy 96 protocol for animal tissues (Qiagen).
The clean DNA in $100\upmu$L elution buffer of 10mM Tris-HCl, 0.5mM EDTA
at pH 9.0 was quantified using Quan-iT PicoGreen dsDNA Assay Kit
(Molecular Probes, Inc.)
and normalized to 5ng/$\upmu$L in nuclease-free water (Qiagen)
for subsequent 16S rRNA illumina sequencing.
16S rRNA amplicon library preparation was performed following a dual-index
paired-end approach \cite{Kozich2013}.
Briefly, PCR amplicon libraries of V4 regions of the 16S rRNA were prepared 
sing dual-index primers (F515/R805), then pooled and sequenced
using the Illumina MiSeq chemistry and platform.
Each sample went through a 30-cycle PCR in duplicate of $20\upmu$L
reaction volumes using 5ng of DNA each, dual index primers, and AccuPrime Pfx SuperMix (Invitrogen).
The thermocycling procedure includes a 2min initial denaturation step at
$95^\circ$C, and 30 cycles of the following PCR scheme:
(a) 20-second denaturation at $95^\circ$C,
(b) 15-second annealing at $55^\circ$C,
and (c) 5-minute extension at $72^\circ$C.
The duplicate PCR products of each sample were pooled, purified, and normalized
using SequalPrep PCR cleanup and normalization kit (Invitrogen).
Barcoded amplicon libraries were then pooled and sequenced using
Illumina Miseq v2 reagent kit, which generated 2$\times$250bp paired-end reads
at the Yale Center for Genome Analysis (YCGA).
The sequencing reads were demultiplexed on QIIME 1.9.0 \cite{Caporaso2010}.
The barcodes, indexes, and primers were removed from raw reads,
producing FASTQ files with both the forward and reverse reads for each sample,
ready for DADA2 analysis \cite{Callahan2017}.
DADA2 version 1.1.6 was used to infer unique biological exact sequence variants
(ESVs) for each sample
and na{\"i}ve Bayes was used to assign taxonomy using the SILVA version 123
database \cite{Wang2007,Quast2013}.

@@

We used the Community Simulator package \cite{Marsland2020} and included new
features for our simulations. In the package,
species are characterized by their resource uptake rates ($c_{i\alpha}$ for
species $i$ and resource $\alpha$), and they all
share a common metabolic matrix $\mathbf{D}$.
The element $D_{\alpha\beta}$
of this matrix represents the fraction of energy in the form of resource $\alpha$
secreted when resource $\beta$ is consumed.
Here we implemented a new operation mode
in which species can secrete different metabolites (and/or
in different abundances) when consuming a same resource. Experimental observations
support the idea of distinct species producing different sets of byproducts when
growing in the same primary resource \mr. We call $D_{i\alpha\beta}$ to the
fraction of energy in the form of resource $\alpha$ secreted \textit{by species
i} when consuming resource $\beta$ ---note that now $D_{i\alpha\beta}$ need not be
equal to $D_{j\alpha\beta}$ if $i \neq j$, unlike in the original Community
Simulator. In the package's underlying Microbial Consumer Resource Model
\cite{Goldford2018,Marsland2019}, this just means that the energy flux
$J^{\mathrm{out}}_{i\beta}$ now takes the form

@@

For our simulations,
we first generate a library of 660 species (divided into three specialist
families of 200 members each
and a generalist family of 60 members)
and 30 resources (divided into three classes of 10 members each).
We split this library into two non-overlapping pools of 330 species each.
We randomly sample 50 species from each pool in equal ratios to seed
100 resident and
100 invasive communities respectively.
We then grow and dilute the communities serially,
replenishing the primary
resource after each dilution.
We repeat the process 20 times to ensure generational equilibrium is
achieved \cite{Goldford2018}.
We then perform the \textit{in silico} experiments by using the
generationally stable communities to seed 100 coalesced communities
that we again stabilize as described previously.
Similarly, we identify the dominant (most
abundant) species of every resident and invasive community to carry out pairwise
competition and single invasion simulations.
Most parameters are set to the defaults of the original Community Simulator
package. Table \mr shows those that are given non-default values to ensure
enough variation in the primary communities.

@@

Experimental data and code for the analysis, as well as code for the simulations
and the updated Community Simulator package with instructions for the
new features can be found in \url{github.com/jdiazc9/coalescence}.

@@

\section*{Test section}

>>>>

% title, authors, affiliations
\title{Cohesiveness in microbial community coalescence}
\author[1,2,3]{Authors}
\author[1,2,$\dagger$]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology,
Yale University, New Haven, CT, USA}
\affil[2]{Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[3]{Other affiliations...}
\affil[$\dagger$]{To whom correspondence should be addressed: \normalfont alvaro.sanchez@yale.edu}
\date{}

@@

Microbial communities often invade one another.
This has been observed, for instance, 
in river courses where terrestrial microbial communities mix with aquatic microorganisms
\cite{Mansour2018,Luo2020,Vass2021}
or in soil communities being invaded as a result of
tillage and outplanting \cite{Rillig2016} or by
aerially dispersed bacteria and funghi \cite{Evans2019}.
Gut microbiomes can invade external communities
through the host animal secretions \cite{Dutton2021},
and the skin microbiota is also subject to
invasions when making contact with environmental sources of microbes \cite{Vandegrift2019}.

The phenomenon by which entire microbiomes invade one another has been termed
\textit{community coalescence} \cite{Rillig2015}.
Ecologists have long contemplated the idea that interactions between multiple co-invading species
can produce correlated invasional outcomes
\cite{Gilpin1994,Simberloff1999,Grosholz2005,Simberloff2006,Gurevitch2006,Green2011,
Livingston2013,Prior2015,Rillig2015,OLoughlin2017,Castledine2020}.
\iffalse
One example is the hypothesis known as \textit{invasional meltdown}, which proposes that
positive interactions between co-invading species can enhance their invasive success and
facilitate future invasions \cite{Simberloff1999,Simberloff2006,Gurevitch2006}.
\fi
However, and
in spite of its clear potential importance, the role of coalescence in microbiome assembly is
only beginning to be addressed and little is known about the mechanisms that govern it and its
potential implications.
Early mathematical models of community-community invasions \cite{Gilpin1994,Toquenaga1997}
as well as more recent work
\cite{Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021}
suggest that high-order invasion effects are common
during community coalescence. Communities that have a previous history of coexistence may exhibit an
emergent ``cohesiveness'' which produces correlated invasional outcomes among species from the
same community \cite{Livingston2013,Sierocinski2017}.
The situation where ecological partners in the invading community recruit each other into the final
coalesced community has been called \textit{ecological co-selection} \cite{Rillig2017,Sierocinski2017}.

The mechanisms of ecological co-selection during community coalescence are still poorly understood.
Do a few key species recruit everyone else, or are collective interactions among all species
(including the rarer members of the community) relevant for coalescence outcomes?
While it is reasonable to expect species with larger population sizes to have a proportionally oversized effect,
natural communities tend to be highly diverse \cite{Louca2016}
and
the role played by the less abundant species has long been subject to debate \cite{Winfree2015}.
Laboratory cultures have also been found to contain uneven distributions of multiple strains that feed
off the metabolic secretions of the dominant species \cite{Rosenzweig1994,Goldford2018}.
The fate of these sub-dominant taxa may be dependent on the invasional success of their dominant
species, or, alternatively, the dominant itself may owe its dominance (at least in part) to cross-feeding
or other forms of facilitation from the rarer members of the population.
These scenarios would give rise to ``top-down'' or ``bottom-up'' community cohesiveness, respectively.
Either of these forms of co-selection could, in principle, be positive (recruitment) or negative (antagonism),
as illustrated in \figref[e]{fig1}.
Which of these situations are typically found in nature?
Previous theoretical and computational studies suggest that the answer is determined by the type and
strength of the interactions of the community members with one another and with the environment
\cite{Tikhonov2016,Vila2019,Lechon2021},
but addressing this question has been experimentally challenging in the past \cite{Rillig2017,Sierocinski2017}.

In previous work, we have shown that a large amount of soil and plant microbiomes can be cultured
\textit{ex situ} in synthetic minimal environments with a single supplied limiting resource under serial
growth-dilution cycles \cite{Goldford2018} (\figref[a-b]{fig1}). Under these conditions, environmental
microbiomes spontaneously re-assemble into complex multi-species communities sustained by dense
cross-feeding facilitation networks \cite{Goldford2018}. In addition,
and just like in natural consortia,
species abundance distributions
in these communities are generally long-tailed and uneven (\figref[d]{fig1} and \figref{figS1}),
with the dominant (most abundant) species typically comprising most of the biomass ($\text{median}=46\%$,
\figref{figS1}). Because these communities are easy to manipulate and grow in high throughput, and
are largely made up by culturable members, they represent good test cases to investigate ecological
co-selection during community coalescence. Here we focus on the dominants and ask whether they can
co-select or be co-selected by the sub-dominant species in their communities (henceforth referred to as
their \textit{cohorts}, \figref[c]{fig1}).

Our results indicate that co-selection varies in direction and strength depending on the supplied limiting
resource.
This primary resource, in turn, has been shown to
shape the structure and interactions of the communities \cite{Estrela2020}.
We observe that, when top-down co-selection is weak, bottom-up co-selection can be very strong,
with positive co-selection being far more common than negative co-selection.
\iffalse
Our results confirm that positive top-down co-selection is common, but its effects are weak.
In contrast, bottom-up co-selection can be very strong, and positive co-selection is far more common
than negative co-selection.
\fi
We then turn to a microbial consumer-resource model (microCRM)
\cite{Goldford2018,Marsland2019,Marsland2020}
that is able to capture the dynamics of microbial communities dominated by metabolic interactions,
as is the case for the ones assembled in our experimental conditions \cite{Goldford2018,Estrela2020}.
We show that the empirically observed trends in ecological co-selection are reproduced
with minimal model assumptions, and that
tuning the complexity of the metabolic interactions in our \textit{in silico} communities can modulate
the recurrence of top-down or bottom-up co-selection.
Our findings indicate that collective interactions play an important role at dictating community structure
during coalescence.

@@

We collected eight natural microbiomes from different soil and plant environmental samples
(\figref[a]{fig1})
and used them to inoculate our synthetic communities, which were stabilized 
in serial batch-culture biorreactors for 84 generations in synthetic minimal media containing either
glutamine or citrate as the only supplied carbon source
(\figref[b]{fig1}, \methodsref{community-assembly}).
We chose these two carbon sources because they are metabolized through different
pathways in bacteria \cite{Dimroth2004,Forchhammer2007},
and we hypothesize that communities assembled in either resource will be supported by
cross-feeding networks of distinct sets of metabolites \cite{Goldford2018,Estrela2020}
thus leading to potentially variable degrees of community cohesiveness
and coalescence outcomes \cite{Tikhonov2016,Tikhonov2017,Castledine2020,Lechon2021}.
We isolated the dominant species of every community
(\methodsref{dominants})
and identified them by Sanger-sequencing their 16S rRNA gene
(\methodsref{sequencing}), which correctly matched the dominant
Exact Sequence Variant (ESV) \cite{Callahan2016,Callahan2017} found through
community-level 16S Illumina sequencing
(\figref{figS1}).
These dominants remained at high frequency after seven additional transfers
with the exception of two of the citrate communities and one of the glutamine
communities (where the dominants were presumably a transiently dominating species)
that were excluded from further analysis
(\figref{figS1}).
Similarly, pairs of communities where the dominants shared a same 16S sequence
and had similar colony morphology were excluded
(\figref{figS1}).

\subsubsection*{Top-down ecological co-selection}

If communities being coalesced were highly cohesive from the top-down,
the dominant species would co-select the rarer members of its community during coalescence
(\figref[e]{fig1}, left panels).
In this scenario, we would expect the outcome of community coalescence to be predicted by
which of the two dominants is most competitive in pairwise competition.
To test this hypothesis, we performed all pairwise competitions between dominant species
in glutamine and citrate environments by mixing them 1:1 on their native media and propagating
the cultures for seven serial transfers, roughly 42 generations
(\methodsref{competitions}).
We then performed all possible pairwise community coalescence experiments by mixing
equal volumes of the communities and propagating the resulting cultures for seven extra
transfers
(\figref[f]{fig1}),
and we quantified coalescence outcomes using the similarity between the coalesced and the
invasive communities
(\methodsref{metrics}).
The frequencies of all species in both community-community and dominant-dominant competitions
were determined by 16S Illumina sequencing
(\methodsref{sequencing}).

We found that, for communities assembled in the glutamine environment,
the pairwise competitive ability of an invasive dominant
is only weakly predictive of the performance of the invasive community in coalescence
(as quantified by the relative Bray-Curtis similarity between the coalesced and invasive communities),
but this correlation was stronger for the citrate communities
(\figref[a]{fig2}, $R^2=0.15$, $p<0.05$ for glutamine and $R^2=0.57$, $p<10^{-4}$ for citrate).
Alternative quantifications of community distance yielded similar results,
with weaker effects when the metric used
accounted only for the presence/absence of specific species and not for their relative abundance in
the communities
(\figref{figS2}).
All these metrics include the presence of the dominant species themselves. To better disentangle the effect
that these dominants have on the other members of their communities, we repeated the analysis this time
excluding the dominant species from the compositional data, finding that our results still hold
(\figref{figS3}).
Furthermore, we observed that, in the glutamine communities,
the relative frequency of a dominant against another in head-to-head pairwise competition
is barely predictive of its relative frequency against that same other dominant when the
cohorts are present too, i.e. after coalescence
(\figref{figS4}, $R^2=0.04$, $p>0.05$).
On the other hand, pairwise competition between dominants of the citrate communities
is highly correlated with the relative abundance of said dominants in community coalescence
 (\figref{figS4}, $R^2=0.83$, $p<10^{-8}$).

Together, these results suggest that the strength of top-down co-selection depends on
the primary resource supplied to the coalesced communities.
Communities assembled with citrate as the primary supplied resource
display a strong degree of top-down co-selection,
with the fates of the sub-dominant species determined to a large extent by dominant-dominant
pairwise competition. This competition is, in turn, only weakly affected by the presence of the cohorts.
For glutamine communities, although some level of top-down co-selection is consistent with our data,
the cohorts do not appear to be passively responding to their dominants.
Instead, evidence suggests that the cohorts might be playing an active role in community coalescence.
This finding led us to investigate the potential role of bottom-up ecological co-selection
(\figref[e]{fig1}, right panels), i.e. whether in some cases the dominants may be co-selected
for or against by their cohorts.

\subsubsection*{Bottom-up co-selection during community coalescence}

To study the effect of bottom-up co-selection, we compared the invasion success of the dominant
from one of the communities when invading the other community in isolation versus when doing so
accompanied by its cohort.
In situations where positive bottom-up ecological co-selection is strong, we expect to see dominants
invading more effectively with their cohorts than by themselves
(\figref[b]{fig2}, green shaded region).
On the other hand, a high degree of negative bottom-up co-selection would result in dominants reaching
higher invasion success alone than in the presence of their cohorts
(\figref[b]{fig2}, red shaded region).
Alternatively, if both forms of bottom-up co-selection were weak, we would see a similar invasion
success of the dominant regardless of whether it is invading with its cohort
(\figref[b]{fig2}, gray shaded region).

We performed a new round of invasion experiments where each dominant invaded every
other community in isolation (\methodsref{competitions}). The dominant-invaded
communities were then propagated for seven serial transfers, and the final
community compositions were determined by 16S Illumina sequencing.
We compared the relative abundances reached by the dominants in these experiments
with the relative abundances of the same dominants in community coalescence
(i.e. when they invade with their cohorts).
\figref[b]{fig2} shows that, for communities assembled in citrate environments,
the cohorts had little effect and the dominants were able to invade with comparable
success regardless of being accompanied by them.
This was expected since we had seen that coalescence between citrate communities
tended to be determined by top-down co-selection.
However, glutamine communities displayed a different behavior.
A large number of dominants (roughly 50\%) that were not able to invade by themselves
(or could only do so at very low final relative abundances, below 0.1)
were able to reach much higher frequencies when they invaded together with their cohorts,
which indicates that positive bottom-up co-selection is common in our experiments.
In contrast, negative co-selection appears to be rare and no instances of it were
observed in our experiments.

\subsubsection*{Simulations of community coalescence with a consumer-resource model}

We ran a set of simulations of community coalescence to better understand the mechanisms
of ecological co-selection in our experiments. We used a microbial consumer-resource
model (microCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
for Python \cite{Marsland2020}. We chose this modeling framework because
communities assembled under our experimental conditions (natural microbiomes re-assembled into
multispecies communities through serial growth-dilution cycles in synthetic minimal media
with a single carbon source)
have been shown to be
sustained by dense metabolic cross-feeding networks \cite{Goldford2018}, for which the
microCRM provides a good description.
We implemented a new feature in the package that makes individual species
able to secrete different sets of byproducts when consuming a same resource
(\methodsref{sim}).
We also used sparse metabolic matrices, equivalent to assuming that species secrete only a few
types of byproducts when consuming a resource. These assumptions are supported by the experimental
observation that individual species can indeed secrete different sets of few metabolites
to the environment when consuming a primary resource
\cite{Harcombe2014,Pinu2018}.
We reason that these choices are necessary to potentially generate community cohesiveness:
if the secretions of all species were roughly similar, the effect of interchanging members of
a community would be minimal. But in practice, species with a history of coexistence make up
cohesive communities with highly specific cross-feeding structures \cite{Goldford2018,Estrela2020}.
In short,
we replicated our experimental protocol \textit{in silico}
by generating a library of species that we divided into two pools. Each pool was used to seed a
collection of invasive and resident communities, that were then allowed to stabilize through 20
growth-dilution cycles. Stable communities were then used to reproduce our coalescence
and dominant-dominant competition experiments,
as well as the invasion experiments from the dominants alone.
A full description of our modeling framework can be found in the \methodsref{sim} section.

We carried out two sets of simulations under two different regimes.
In one case, we generated a library of 30 resources divided into 3 resource classes
(\methodsref{sim}). This corresponds to a situation where the digestion of a resource
can yield a wide variety of secreted byproducts.
We refer to this scenario as the \textit{complex resource} regime.
Our second set of simulations was performed under a \textit{simple resource} regime, where we
generated a library of only 15 resources belonging to a same class.
For the simple resource, we found that coalescence outcomes were highly correlated with
the competitive ability of the community dominants in pairwise competition
(\figref[c]{fig2}),
indicating a large degree of top-down positive co-selection.
These dominants were able to reach similar fractions in the final communities when invading alone
with respect to when invading with their cohorts
(\figref[d]{fig2}),
suggesting that bottom-up co-selection is weak in this scenario.
On the other hand,
simulations under the complex resource regime evidenced weak top-down co-selection but 
common and strong bottom-up recruitment (\figref[e-f]{fig2}), with negative co-selection remaining rare.
Simulations suggest that community cohesiveness is dependent on the metabolic interactions among species,
which is in line with previous findings \cite{Goldford2018, Estrela2020}.
These interactions are determined by the metabolic profile of the primary resource in the conditions
of our simulations and, in turn, modulate the strength and direction of ecological co-selection.

\subsubsection*{Conclusion}

Understanding the mechanisms underlying the responses of microbial communities to invasions is
an essential but poorly understood question in microbial ecology \cite{Rillig2015}.
Theory has suggested that communities may exhibit an emergent cohesiveness
\cite{Gilpin1994,Livingston2013,Tikhonov2016,Tikhonov2017},
leading to members of the same community recruiting one another during
community-community invasions.
Our results provide direct experimental evidence of ecological co-selection in a large number
of community coalescence experiments,
and highlight the critical role played by the rarer, sub-dominant species in the generation
of community cohesiveness.
Our data suggests that the strength and direction of ecological co-selection is modulated
by the supplied primary carbon source,
which in turn shapes the cross-feeding network that dictates
the structure of our communities \cite{Goldford2018,Estrela2020}.
These observations,
together with previous experimental results in different systems \cite{Sierocinski2017} as well as
theoretical predictions
\cite{Gilpin1994,Toquenaga1997,Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021},
suggest that collective interactions arising from metabolic feedbacks between microbes
and the environment should be generically expected to produce ecological co-selection
during community coalescence.

Further work is required to clarify the relationship between metabolic interactions,
community structure, cohesiveness and co-selection.
The experimental system that we introduced in this work can be easily expanded so that
large numbers of community coalescence experiments can be carried out in parallel.
It thus represents a promising tool to explore the properties of microbial community
coalescence in high throughput
and test quantitative theories about its role in microbiome assembly.

@@

\subsubsection*{Coalescence, competition and invasion experiments}
\label{methods:competitions}

@@

The sequencing protocol was identical to that described in \cite{Goldford2018}.
Community samples were collected by spinning down at 3500rpm for 25min
in a bench-top centrifuge at room temperature;
cell pellets were stored at $-80^\circ$C before processing.
To maximize Gram-positive bacteria cell wall lysis,
the cell pellets were re-suspended and incubated at $37^\circ$C for 30min
in enzymatic lysis buffer (20mM Tris-HCl, 2mM sodium EDTA, 1.2\% Triton X-100)
and 20mg/mL of lysozyme from chicken egg white (Sigma-Aldrich).
After cell lysis, the DNA extraction and purification was performed using the
DNeasy 96 protocol for animal tissues (Qiagen).
The clean DNA in $100\upmu$L elution buffer of 10mM Tris-HCl, 0.5mM EDTA
at pH 9.0 was quantified using Quan-iT PicoGreen dsDNA Assay Kit
(Molecular Probes, Inc.)
and normalized to 5ng/$\upmu$L in nuclease-free water (Qiagen)
for subsequent 16S rRNA Illumina sequencing.
16S rRNA amplicon library preparation was performed following a dual-index
paired-end approach \cite{Kozich2013}.
Briefly, PCR amplicon libraries of V4 regions of the 16S rRNA were prepared 
sing dual-index primers (F515/R805), then pooled and sequenced
using the Illumina MiSeq chemistry and platform.
Each sample went through a 30-cycle PCR in duplicate of $20\upmu$L
reaction volumes using 5ng of DNA each, dual index primers, and AccuPrime Pfx SuperMix (Invitrogen).
The thermocycling procedure includes a 2min initial denaturation step at
$95^\circ$C, and 30 cycles of the following PCR scheme:
(a) 20-second denaturation at $95^\circ$C,
(b) 15-second annealing at $55^\circ$C,
and (c) 5-minute extension at $72^\circ$C.
The duplicate PCR products of each sample were pooled, purified, and normalized
using SequalPrep PCR cleanup and normalization kit (Invitrogen).
Barcoded amplicon libraries were then pooled and sequenced using
Illumina Miseq v2 reagent kit, which generated 2$\times$250bp paired-end reads
at the Yale Center for Genome Analysis (YCGA).
The sequencing reads were demultiplexed on QIIME 1.9.0 \cite{Caporaso2010}.
The barcodes, indexes, and primers were removed from raw reads,
producing FASTQ files with both the forward and reverse reads for each sample,
ready for DADA2 analysis \cite{Callahan2017}.
DADA2 version 1.1.6 was used to infer unique biological exact sequence variants
(ESVs) for each sample
and na{\"i}ve Bayes was used to assign taxonomy using the SILVA version 123
database \cite{Wang2007,Quast2013}.

@@

We used the Community Simulator package \cite{Marsland2020} and included new
features for our simulations. In the package,
species are characterized by their resource uptake rates ($c_{i\alpha}$ for
species $i$ and resource $\alpha$), and they all
share a common metabolic matrix $\mathbf{D}$.
The element $D_{\alpha\beta}$
of this matrix represents the fraction of energy in the form of resource $\alpha$
secreted when resource $\beta$ is consumed.
Here we implemented a new operation mode
in which species can secrete different metabolites (and/or
in different abundances) when consuming a same resource. Experimental observations
support the idea of distinct species producing different sets of byproducts when
feeding off the same primary resource
\cite{Harcombe2014,Pinu2018}.
We call $D_{i\alpha\beta}$ to the
fraction of energy in the form of resource $\alpha$ secreted \textit{by species
i} when consuming resource $\beta$ ---note that now $D_{i\alpha\beta}$ need not be
equal to $D_{j\alpha\beta}$ if $i \neq j$, unlike in the original Community
Simulator. In the package's underlying Microbial Consumer Resource Model
\cite{Goldford2018,Marsland2019}, this just means that the energy flux
$J^{\mathrm{out}}_{i\beta}$ now takes the form

@@

For our simulations,
we first generated a library of 2400 species divided into three specialist
families of 800 members each
and a generalist family of 240 members.
We split this library into two non-overlapping pools of 1320 species each.
We randomly sample 50 species from each pool in equal ratios to seed
100 resident and
100 invasive communities respectively.
We then let grow and diluted the communities serially,
replenishing the primary
resource after each dilution.
We repeated the process 20 times to ensure generational equilibrium is
achieved \cite{Goldford2018}.
We then performed the \textit{in silico} experiments by using the
generationally stable communities to seed 100 coalesced communities
that we again stabilize as described previously.
Similarly, we identified the dominant (most
abundant) species of every resident and invasive community to carry out pairwise
competition and single invasion simulations.

Two sets of simulations were carried out, corresponding to a
\textit{simple} or \textit{complex} externally replenished resource respectively
(see details in main text).
For the \textit{simple} resource scenario, we generated a library of 15 resources
belonging to a same class. For the \textit{complex} resource scenario,
the library contained 30 resources subdivided into 3 classes
(each being preferentially consumed by one of the three species families).
Our aim with this approach was to examine the implications of allowing for
different degrees of density in the metabolic networks sustaining our communities.
Most other parameters were set to the defaults of the original Community Simulator
package, with the only exception of the maintenance costs ($m$) which are set to
zero for all species (as cell death is negligible through the duration of our growth cycles)
and the sparsity of the metabolic matrices ($s$) which is set to 0.9
to generate significant variability in the secretion fluxes across different species
(see main text).

@@

Experimental data and code for the analysis, as well as code for the simulations
and the updated Community Simulator package with instructions for the
new features are deposited in \url{github.com/jdiazc9/coalescence}.

@@
